Information Closure Theory of Consciousness

### 2020-03-10 ICT_Frontiers_rev202001.tex
Move to local writing

\ProvidesPackage{WritingToolsByAcer}

@@

%% Title style
\usepackage{stringstrings}
\usepackage[explicit]{titlesec}
\usepackage{sectsty}

@@

		It seems to be true that we are not consciously accessing information processed at every scale in the neural system. There are both more microscopic and more macroscopic levels than the level corresponding to the conscious contents. On the one hand, dynamics of individual neurons are stochastic \citep{Goldwyn2011, White2000}. However, what we are aware of in our conscious mind shows astonishing stability and robustness against the ubiquitous noise in the neural system \citep{mathis1995computational}. In addition, some parts of the neural system contribute very little to conscious experience (the cerebellum for example \citep{lemon2010life}), also suggesting that conscious contents do not have one-to-one mapping to the entire state of the neural system. On the other hand, human conscious experience is more detailed than just a simple (e.g. binary) process can represent, suggesting that the state space of conscious experience is much larger than what a single overly coarse-grained binary variable can represent. These facts suggest that conscious processes occur at a particular scale. We currently lack a theory to identify the scale which conscious processes correspond to. We refer to this notion as \textbf{the scale problem of consciousness} (Fig.~\ref{fig:scaleproblem}).

@@

		% Definition of informational closure
			Consider two processes, the environment process $(E_t)_{t \in \mathbb{N}}$ and the system's process $(Y_t)_{t \in \mathbb{N}}$ and let their interaction be described by the Bayesian network in Fig.~\ref{fig:SystemAndEnv}. Then, information flow $J_{t}$ from the environment $E$ to a system $S$ at time $t$ can be defined as the conditional mutual information $I$ between the current environment state $E_{t}$  and the future system state $Y_{t+1}$ given the current system state $Y_{t}$

@@

			\noindent
			And, non-trivial informational closure can be defined as
				\begin{align}
				\label{eq:NTIC}
    				% \left.\begin{array}
    				NTIC_t(E\rightarrow Y) :&=I(Y_{t+1};Y_{t})-I(Y_{t+1};Y_{t}|E_{t})\\
    				&=I(Y_{t+1};E_{t})-I(Y_{t+1};E_{t}|Y_{t})
    				% \end{array} \right.
				\end{align}

@@

	   	\begin{definition}
	   	Given a stochastic process $X$ with state space $\mathcal{X}$, a \emph{coarse-graining of $X$} is a stochastic process $Y$ with state space $\mathcal{Y}$ such that there exists a function\footnote{Functions in the mathematical sense used here are always either one-to-one or many-to-one.} $f_Y:\mathcal{X}\rightarrow \mathcal{Y}$ with $Y_t=f_Y(X_t)$.  
	   	\end{definition}
	   	A more general definition of coarse-grainings that maps temporally extended sequences of the microscopic process to macroscopic states are possible but for his first exposure of our theory the simpler definition above is sufficient.

@@

        A concrete example in the context of neuroscience is that $X$ represents the microscopic level of the universe, $S$ a cellular level process in the neural system, $Y$ a more macroscopic process of the neural system coarse-grained from the cellular level process $S$, and $E$ the environment which the cellular level process $S$ interacts with. The environment $E$ may include other processes in the neural system, the sensors for perception and interoception, and external physical worlds.\\

@@

    		Importantly, NTIC processes encodes environmental information in its state. This suggests that a NTIC process can be considered as a process that simulates the environmental dynamics. This implication fits well with some theories of consciousness (for example, world simulation metaphor ~\citep{revonsuo2006inner}). Note that \ac{OurTheory} doesn't assume that generative models are necessary for consciousness. The implication is a natural result of processes with NTIC.

@@

        		The same principle can be applied to interpret blindsight \citep{humphrey1999history, humphrey1974vision, Humphrey1970} and procedural memory \citep{doyon2009contributions, ashby2010cortical} which are often considered as unconscious processes. Blindsight patients are able to track objects, avoid obstacles, and make above chance-level visual judgements without having any conscious perception about visual stimuli. We argue that action outputs of blindsight can be directly guided by sensory inputs through stimulus-response maps. The neural circuits are not informationally closed and, therefore, unconscious. Similarly, for procedural memory, the state transitions of action controls largely depends on the concurrent environmental states. This prevents the internal processes of procedure memory from informational closure and being conscious. \ac{OurTheory} also offer an interpretation as to why patients with visual apperceptive agnosia \citep{james2003ventral} can perform online motor controls without visual awareness of action targets \citep{10.3389/fneur.2014.00255}.

@@

		    \textbf{Prediction after system damaged}: \ac{OurTheory} and IIT lead to different predictions when a system suffers from a damage. For example, considering a densely connected networks whose dynamics forms an NTIC process. If we cut the network in half, IIT would predict that this results in two consciousnesses because elements in both networks still maintain high degrees of interactions. In contrast, \ac{OurTheory} would predict that this operation could completely destroy NTIC rendering both parts unconscious.

@@

    		% compatible things
    		\ac{OurTheory} is well compatible with PP. Crucially, \ac{OurTheory} further provides natural and fundamental explanations to fills the two explanatory gaps which PP encounters. According to the definition of NTIC, a process with high NTIC can be regarded as a powerful predictive machine which has accurate self-predictive information ($I(Y_{t+1};Y_{t})$, E.q.~\ref{eq:NTIC}) and concurrently incorporates environmental information into its dynamic ($I(Y_{t+1};Y_{t}|E_{t})$, E.q.~\ref{eq:NTIC}). This predictive nature of NTIC processes is in agreement with the core notion of PP in which the conscious contents are always the predicted (inferred) outcome of our predictive mechanisms. Second, due to the informational closure to the environment, the encoded information about its environment in an NTIC process can be seemed as "the best guess" about the external environment in the context of Bayesian inference.

@@

@@

        Finally another empirical challenge to \ac{OurTheory} is the lack of empirical supporting evidence. This is understandable because the concept of NTIC is relatively new in the history of information science, not to mention in neuroscience. Very few experiments and data collections are designed for examining NTIC properties in neural systems. To our knowledge, only two studies \citep{Palmer2015, sederberg2018learning} coincidentally examine relevant properties in salamander retina. They found that the a large group of neural populations of retinal ganglion cells encoded predictive information about external stimuli also had high self-predictive information about their own future states. This result is in line with the characteristic of NTIC. We expect that there will be more empirical studies examining relevant neural properties of NTIC.

>>>>

%\ProvidesPackage{WritingToolsByAcer}

@@

%% Title style
\usepackage{stringstrings}
\usepackage[explicit]{titlesec}
%\usepackage{sectsty}

@@

		It seems to be true that we are not consciously accessing information processed at every scale in the neural system. There are both more microscopic and more macroscopic levels than the level corresponding to the conscious contents. On the one hand, dynamics of individual neurons are stochastic \citep{Goldwyn2011, White2000}. However, what we are aware of in our conscious mind shows astonishing stability and robustness against the ubiquitous noise in the neural system \citep{mathis1995computational}. In addition, some parts of the neural system contribute very little to conscious experience (the cerebellum for example \citep{lemon2010life}), also suggesting that conscious contents do not have one-to-one mapping to the entire state of the neural system. On the other hand, human conscious experience is more detailed than just a simple (e.g. binary) process can represent, suggesting that the state space of conscious experience is much larger than what a single overly coarse-grained binary variable can represent. These facts suggest that conscious processes occur at a particular scale. \linelabel{line:lack-theory}We currently lack a theory to identify the scale which conscious processes correspond to. We refer to this notion as \textbf{the scale problem of consciousness} (Fig.~\ref{fig:scaleproblem}).

@@

		% Definition of informational closure
			\linelabel{line:r1q1a}Consider two processes, the environment process $(E_t)_{t \in \mathbb{N}}$ and the system's process $(Y_t)_{t \in \mathbb{N}}$ and let their interaction be described by the Bayesian network in Fig.~\ref{fig:SystemAndEnv}. Then, information flow $J_{t}$ from the environment $E$ to a system $S$ at time $t$ can be defined as the conditional mutual information $I$ between the current environment state $E_{t}$  and the future system state $Y_{t+1}$ given the current system state $Y_{t}$

@@

			\noindent
			And, non-trivial informational closure can be defined as
				\begin{align}
%				\label{eq:NTIC}
    				% \left.\begin{array}
    				NTIC_t(E\rightarrow Y) :&=I(Y_{t+1};Y_{t})-I(Y_{t+1};Y_{t}|E_{t}) 
    				\label{eq:NTIC1}\\
    				&=I(Y_{t+1};E_{t})-I(Y_{t+1};E_{t}|Y_{t}) 
    				\label{eq:NTIC2}
    				% \end{array} \right.
				\end{align}

@@

	   	\begin{definition}
	   	\linelabel{line:r2q1-1} Given a stochastic process $X$ with state space $\mathcal{X}$, a \emph{coarse-graining of $X$} is a stochastic process $Y$ with state space $\mathcal{Y}$ such that there exists a function \footnote{Functions in the mathematical sense used here are always either one-to-one or many-to-one.} $f_Y:\mathcal{X}\rightarrow \mathcal{Y}$ with $Y_t=f_Y(X_t)$.  
	   	\end{definition}
	   	A more general definition of coarse-grainings that maps temporally extended sequences of the microscopic process to macroscopic states are possible but for his first exposure of our theory the simpler definition above is sufficient.

@@

        A concrete example in the context of neuroscience is that $X$ represents the microscopic level of the universe, $S$ a cellular level process in the neural system, $Y$ a more macroscopic process of the neural system coarse-grained from the cellular level process $S$, and $E$ the environment which the cellular level process $S$ interacts with. \linelabel{line:neu-env} The environment $E$ may include other processes in the neural system, the sensors for perception and interoception, and external physical worlds.\\

@@

    		Importantly, NTIC processes encode environmental information in its state. This suggests that a NTIC process can be considered as a process that simulates the environmental dynamics. This implication fits well with some theories of consciousness (for example, world simulation metaphor ~\citep{revonsuo2006inner}). Note that \ac{OurTheory} doesn't assume that generative models are necessary for consciousness. The implication is a natural result of processes with NTIC.

@@

        		The same principle can be applied to interpret blindsight \citep{humphrey1999history, humphrey1974vision, Humphrey1970} and procedural memory \citep{doyon2009contributions, ashby2010cortical} which are often considered as unconscious processes.
        		\linelabel{line:r2q6}Blindsight patients are able to track objects, avoid obstacles, and make above chance-level visual judgements when their visual experiences is degraded or missing (however, they may still preserve some forms of conscious experience, see \cite{overgaard2011visual, mazzi2016blind}). We argue that action outputs of blindsight can be directly guided by sensory inputs through stimulus-response maps. The neural circuits are not informationally closed and, therefore, unconscious. Similarly, for procedural memory, the state transitions of action controls largely depends on the concurrent environmental states. This prevents the internal processes of procedure memory from informational closure and being conscious. \ac{OurTheory} also offer an interpretation as to why patients with visual apperceptive agnosia \citep{james2003ventral} can perform online motor controls without visual awareness of action targets \citep{10.3389/fneur.2014.00255}.

@@

		    \textbf{Prediction after system damaged}:\linelabel{line:r2-cutting} \ac{OurTheory} and IIT lead to different predictions when a system suffers from a damage. For example, considering a densely connected networks whose dynamics forms an NTIC process. If we cut the network in half, IIT would predict that this results in two consciousnesses because elements in both networks still maintain high degrees of interactions. In contrast, \ac{OurTheory} would predict that this operation could completely destroy NTIC rendering both parts unconscious.

@@

    		% compatible things
    		\ac{OurTheory} is well compatible with PP. Crucially, \ac{OurTheory} further provides natural and fundamental explanations to fills the two explanatory gaps which PP encounters. According to the definition of NTIC, a process with high NTIC can be regarded as a powerful predictive machine which has accurate self-predictive information ($I(Y_{t+1};Y_{t})$, E.q.~\ref{eq:NTIC1}) and concurrently incorporates environmental information into its dynamic ($I(Y_{t+1};Y_{t}|E_{t})$, E.q.~\ref{eq:NTIC1}). This predictive nature of NTIC processes is in agreement with the core notion of PP in which the conscious contents are always the predicted (inferred) outcome of our predictive mechanisms. Second, due to the informational closure to the environment, the encoded information about its environment in an NTIC process can be seemed as "the best guess" about the external environment in the context of Bayesian inference.

@@

        \linelabel{line:dream}

@@

        \linelabel{line:r1-state-dep}

        Finally another empirical challenge to \ac{OurTheory} is the  of empirical supporting evidence. This is understandable because the concept of NTIC is relatively new in the history of information science, not to mention in neuroscience. Very few experiments and data collections are designed for examining NTIC properties in neural systems. To our knowledge, only two studies \citep{Palmer2015, sederberg2018learning} coincidentally examine relevant properties in salamander retina. They found that the a large group of neural populations of retinal ganglion cells encoded predictive information about external stimuli also had high self-predictive information about their own future states. This result is in line with the characteristic of NTIC. We expect that there will be more empirical studies examining relevant neural properties of NTIC.
